# Tangential Majorization

## Using the Tangent

### Majorizing and Minorizing the Logarithm

The logarithm is concave. Consequently, for all positive $x$ and $y$, we have the linear majorizer
$$
\log x\leq\log y+\frac{1}{y}(x-y)=\log y+\frac{x}{y}-1.
$$
We can apply the same concavity to get a minorizer
$$
\log\frac{1}{x}\leq\log\frac{1}{y}+y(\frac{1}{x}-\frac{1}{y}),
$$
which is
$$
\log x\geq \log y-\frac{y}{x}+1.
$$
These majorizers and minorizers of the logarithm are illustrated in Figure 1 for $y=1$ and $y=5$.
<hr>

```{r mylog,fig.align="center",echo=FALSE,cache=TRUE}
 x<-seq(.5,10,length=100)
 lx<-log(x)
 u1<-log(1)+(x/1)-1
 u5<-log(5)+(x/5)-1
 l1<-log(1)-(1/x)+1
 l5<-log(5)-(5/x)+1
 plot(x,lx,type="l",col="RED",lwd=3,ylab="log(x)")
 lines(x,u1,col="BLUE",lwd=2)
 lines(x,u5,col="BLUE",lwd=2)
 lines(x,l1,col="GREEN",lwd=2)
 lines(x,l5,col="GREEN",lwd=2)
 abline(v=1,lwd=2)
 abline(v=5,lwd=2)
```
<center>
Figure 1: Majorization and Minorization of Logarithm
</center>
<hr>
More generally, we have
$$
\log x\leq\log y+\frac{1}{p}\left\{\left(\frac{x}{y}\right)^p-1\right\}
$$
for all $p>0$ and
$$
\log x\geq\log y+\frac{1}{p}\left\{\left(\frac{x}{y}\right)^p-1\right\}
$$
for all $p<0$. See Figure 2, where $y=5$ and $p=\{-2,-1,1,2\}$.
<hr>

```{r morelog,fig.align="center",echo=FALSE,cache=TRUE}
 x<-seq(.5,10,length=100)
 lx<-log(x)
 y <- 5
 p <- -2
 u1<-log(y)+(1/p)*((x/y)^p-1)
 p<--1
 u2<-log(y)+(1/p)*((x/y)^p-1)
 p<-1
 l1<-log(y)+(1/p)*((x/y)^p-1)
 p<-2
 l2<-log(y)+(1/p)*((x/y)^p-1)
 plot(x,lx,type="l",col="RED",lwd=3,ylab="log(x)")
 lines(x,u1,col="GREEN",lwd=2)
 lines(x,u2,col="GREEN",lwd=2)
 lines(x,l1,col="BLUE",lwd=2)
 lines(x,l2,col="BLUE",lwd=2)
 abline(v=y,lwd=2)
```
<center>
Figure 2: Majorization and Minorization of Logarithm
</center>
<hr>

### Aspects of Correlation Matrices

Suppose $\underline{x}_j,\cdots,\underline{x}_m$ are random variables, and $\mathcal{K}_j$ are convex cones of Borel-measurable real-valued functions of $\underline{x}_j$ with finite variance. The elements of $\mathcal{K}_j$ are called _transformations_ of the variable $\underline{x}_j$.

For instance, $\mathcal{K}_j$ can be the cone of monotone transformations, or the subspace of splines with given knots, or the subspace of quantifications of a categorical variable

A transformation $\kappa\in\mathcal{K}$ is standardized if $\mathbf{E}(\kappa(\underline{x}))=0$ and $\mathbf{E}(\kappa^2(\underline{x}))=1$. Standardized transformations define a sphere $\mathcal{S}_j$.


Now suppose $f$ is a
concave and differentiable function defined on the space of all correlation matrices $R$ between $m$ random variables. Suppose we want to minimize
$$
g(\kappa_1,\cdots,\kappa_m)\mathop{=}\limits^{\Delta}f(R(\kappa_1(\underline{x}_1),\cdots,\kappa_m(\underline{x}_m)))
$$
over all transformations $\kappa_j\in\mathcal{K}_j\cap\mathcal{S}_j$.

Because $f$ is concave
$$f(R)\leq f(S)+\hbox{ tr }\nabla f(S)(R-S).$$
Collect the partials in the matrix $G.$
A majorization algorithm can minimize
$$
\sum_{i=1}^m\sum_{j=1}^m g_{ij}(S)\mathbf{E}(\kappa_i\kappa_j),
$$
over all standardized transformations, which we do with block relaxation using $m$ blocks. In each block we must maximize a linear function on a cone, under a quadratic constraint, which is usually not hard to do.

This algorithm generalizes ACE, CA, and many other forms of MVA with OS. It was proposed first by [De Leeuw [1988a]](), with additional theretical results in [De Leeuw [1988b]](). The function $f$ can be based
on multiple correlations, eigenvalues, determinants, and so on.

### Partially Observed Linear Systems


@deleeuw_C_04a discusses the problem of finding an approximate solution to the homogeneous linear system $AB=0$ when there are cone and orthonormality restrictions on the columns of $A$ and when some elements of $B$ are restricted to known values, most commonly to zero. Think of the columns of $A$ as variables or sets of variables, and think of $B$ as regression coefficients or weights.

The loss function used by @deleeuw_C_04a is
$$
f(R)\mathop{=}\limits^{\Delta}\min_{B\in\mathcal{B}}\mathbf{tr} B'RB,\tag{1}
$$
with $R\mathop{=}\limits^{\Delta}A'A$ and with $\mathcal{B}$ coding the constraints on $B$. Note that the computation of the optimal $B$ in $\text{(1)}$ is a least squares problem, and even with linear inequality constraints on $B$ it is still a straightforward quadratic programming problem.

The function $f$ in $\text{(1)}$ is the pointwise minimum of linear functions in $R$, and thus it is a concave function of $R$.
This means we are in the "aspects of correlation matrices" framework discussed in the previous section.

In particular we define
$$
B(R)\mathop{=}\limits^{\Delta}i\{\hat B\mid\mathbf{tr}\ \hat B'R\hat B=\min_{B\in\mathcal{B}}\mathbf{tr}\ B'RB\},
$$
then the subgradient of $f$ at $R$ is
$$
\partial f(R)=\mathbf{conv}(BB'\mid B\in B(R)).
$$
The subgradient inequality now says that for all correlation matrices $R$ and $S$ we have $f(R)\leq \mathbf{tr}\ \nabla R$
for all $\nabla\in\partial f(S)$.

The constraints on $A$ discussed in @deleeuw_C_04a make it possible to fit a wide variety of multivariate analysis techniques. Columns of $A$, or variables, are partitioned into blocks. Some blocks contain only a single variable, such variables are called _single_. Some blocks are constrained to be orthoblocks, which means that the variabes in the block are required to be orthonormal. Single variables may be cone-constrained, which means the corresponding column of $A$ is
constrained to be in a cone in $\mathbb{R}^n$. And orthoblocks may be subspace-constrained, which means all columns must be in the same subspace.

We mention some illustrative special cases here. Common factor analysis of a data matrix $Y$ means finding an approximate solution to the system
$$
\begin{bmatrix}
Y&\mid&U&\mid&E
\end{bmatrix}
\begin{bmatrix}
\ \ I\\
-\Gamma\\
-\Delta
\end{bmatrix} = 0
$$
with $U'U=I$, $E'E=I$, $U'E=0$, and $\Delta$ diagonal. The _common factor scores_ are in $U$, the _unique factor scores_ in $E$, the _factor loadings_ in $\Gamma$ and the _uniquenesses_ in $\Delta$. This example can be generalized to cover structural equation models

Homeogeneity analysis @gifi_B_90 is the linear system
$$
\begin{bmatrix}
X&\mid&Q_1&\mid&\cdots&\mid&Q_m
\end{bmatrix}
\begin{bmatrix}
I&I&I&\cdots&I\\
-\Gamma_1&0&0&\cdots&0\\
0&-\Gamma_2&0&\cdots&0\\
0&0&-\Gamma_3&\cdots&0\\
\vdots&\vdots&\vdots&\ddots&\vdots\\
0&0&0&\cdots&-\Gamma_m
\end{bmatrix}=0
$$
where $X$ is and orthoblock of object scores, while the $Q_j$ are orthoblocks in the subspaces defined by the indicator matrices (or B-spline bases) of variable $j$. For single variables $Q_j$ only has a single column, which can be cone-constrained. For _multiple correspondence analysis_ $X$ and all $Q_j$ have the same number of columns. For _nonlinear principal component analysis_ all variables are single and the $\Gamma_j$ are $1\times p$.



In both examples the majorization algorithm is actually an alternating least squares algorithm. In the factor analysis example the loss functon is
$$
\sigma(Y,U,\Gamma,E,\Delta)=\|Y-U\Gamma-E\Delta\|^2,
$$
and in homogeneity analysis it is
$$
\sigma(X,Q_1\cdots,Q_m,\Gamma_1,\cdots,\Gamma_j)=\sum_{j=1}^m\|X-Q_j\Gamma_j\|^2.
$$

### Gpower

** Rewrite for minimizing a concave function 02/21/15 **

Consider the problem of maximizing a convex function $f$ on a compact set $\mathbf{X}$. The function is not necessarily differentiable, the constraint set is not necessarily convex. Define
$$
f^\star\mathop{=}\limits^{\Delta}\max_{x\in\mathcal{X}} f(x).
$$
For all $x,y$ and $z\in\partial f(y)$ we have  the subgradient inequality
$$
f(x)\geq f(y)+z'(x-y).
$$
Thus the majorization algorithm is
$$
x^{(k+1)}\in\mathbf{Arg}\mathop{\mathbf{max}}\limits_{x\in\mathcal{X}} z'x,
$$
where $z$ is any element of $\partial f(x^{(k)})$.

Define
$$
\delta(y)\mathop{=}\limits^{\Delta}\max_{x\in\mathcal{X}} z'(x-y)
$$
where $z\in\partial f(y)$. Then $\delta(y)\geq 0$ and $\delta(y)$
vanishes only when $z$ is in the normal cone to $\mathbf{conv}(\mathcal{X})$ at $y$.
$$
f(x^{(k+1)})\geq f(x^{(k)})+\delta(x^{(k)}).
$$
It follows that
$$
f(x^{(k+1)})-f(x^{(0)})=\sum_{i=0}^k(f(x^{(i+1)})-f(x^{(i)}))\geq\sum_{i=0}^k\delta(x^{(i)}),
$$
and
$$
S_k\mathop{=}\limits^{\Delta}\sum_{i=0}^k\delta(x^{(i)})\leq f^\star-f(x^{(0)}).\tag{1}
$$
Thus the partial sums $S_k$ define an increasing sequence, which is bounded above and consequently converges. This implies its terms converge to zero. i.e.
$$
\lim_{k\rightarrow\infty}\delta(x^{(k)})=0.
$$
If
$$
\delta_k\mathop{=}\limits^{\Delta}\min_{0\leq i\leq k}\delta(x^{(i)}),
$$
then, from $\text{(1)}$,
$$
\delta_k\leq\frac{f^*-f(x^{(0)})}{k+1}.
$$

## Broadening the Scope

### Differences of Convex Functions

For d.c. functions (differences of convex functions) such as $\phi=\alpha-\beta$ we can write
$\phi(\omega)\leq\alpha(\omega)-\beta(\xi)-\eta'(\omega-\xi),$
with $\eta\in\partial\beta(\xi).$
This gives a convex majorizer. Interesting, because basically all twice differentiable functions are d.c.


** Add: convexification 02/21/15 **

Suppose $f(x)=\frac14 x^4-\frac12 x^2$. If we look for a majorization then our first thought is to majorize the first term, because the second is already nicely quadratic. But in this case we proceed the other way around.

In fact, let's consider the more general case $f(x)=\frac14 x^4-h(x)$, where $h$ is convex and
differentiable. Note $f$ is indeed the difference of
two convex functions. We see
that $f'(x)=x^3-h'(x)$ and $f''(x)=3x^2-h''(x)$.

Using tangential majorization for $h$ gives
$$
g(x,y)=\frac14 x^4-h(y)-h'(y)(x-y).
$$
Clearly $g$ is convex in $x$ for every $y$, and since $\mathcal{D}_1g(x,y)=x^3-h'(y)$ we have
$$
x^{(k+1)}=\sqrt[3]{h'(x^{(k)})}.
$$
The iteration radius at a fixed point turns out to be
$$
\kappa(x)=\frac{h''(x)}{f''(x)+h''(x)}=\frac13\frac{h''(x)}{x^2}.
$$

For $h(x)=\frac12 x^2$ we have $h'(x)=x$. Convergence is to $\pm 1$, and thus, using l'Hôpital's rule,
$$
\kappa(1)=\lim_{x\rightarrow 1}\frac{\sqrt[3]{x}-1}{x-1}=\frac13.
$$
For $h(x)=|x|$ we have $h'(x)=\pm 1$ if $x\not= 0$, and thus $x^{(k+1)}=\pm 1.$ The algorithm finishes in a single step, with the correct solution.

### More DC

#### Peaky

$f(x,y)=\frac12x^2y^2$

$$\nabla f(x,y)=\begin{bmatrix}xy^2\\yx^2\end{bmatrix}$$
$$
\nabla^2f(x,y)=\begin{bmatrix}y^2&2xy\\2xy&x^2\end{bmatrix}
$$
Eigenvalues 
$$
\lambda^2-\lambda(x^2+y^2)-3x^2y^2=0
$$
$$
\lambda_{1,2}=\frac12{(x^2+y^2)\pm\frac12\sqrt{(x^2+y^2)^2+12x^2y^2}}
$$
#### Monkey Saddle

$f(x,y)=\frac16x^3 - \frac12 xy^2$

$$
\nabla f(x,y)=\frac12\begin{bmatrix}x^2-y^2\\-2xy\end{bmatrix}$$
$$
\nabla^2f(x,y)=\begin{bmatrix}x&-y\\-y&-x\end{bmatrix}
$$
$\lambda_1+\lambda_2=0$ $\lambda_1\lambda_2=-(x^2+y^2)$
$\lambda_1=\sqrt{x^2+y^2}$
$\lambda_2=-\sqrt{x^2+y^2}$
eigenvalues $\begin{bmatrix}y&\mid&x-\lambda\end{bmatrix}$
ssq $y^2+(x-\lambda)^2=2\lambda(\lambda-x)$
thus
$$
[\nabla^2f(x,y)]_+=\frac12
\begin{bmatrix}\frac{y^2}{\lambda-x}&-y\\-y&(\lambda-x)\end{bmatrix}
$$


### Convexifiable Functions

### Piecewise Linear Majorization
